Relaxed Weak Accelerated Proximal Gradient Method: a Unified Framework for Nesterov's Accelerations

### 2025-09-18 Notebook LaTEX/R-WAPG/2025-06-10-00-paper.tex
>w<, changes

        Finally, since $\mu = 0$, Propisition \ref{prop:r-wapg-momentum-repr} applies with the momentum term simplifies into $\alpha_{k + 1}(\alpha_k^{-1} - 1)$. 

@@

                \caption{A single LASSO experiment results, with $M = 64, N=256$. f}

>>>>

        Finally, since $\mu = 0$, Proposition \ref{prop:r-wapg-momentum-repr} applies with the momentum term simplifies into $\alpha_{k + 1}(\alpha_k^{-1} - 1)$. 

@@

                \caption{A single LASSO experiment results, with $M = 64, N=256$.}

### 2025-09-18 Notebook LaTEX/R-WAPG/presentation-shorter.tex
>w<, changes

% THEMES AND BEAMER SETTINGS ===================================================
% \usetheme{Madrid}
\graphicspath{{.}}
\input{presets/wang/custom_commands.tex} % IMPORT WANG'S LATEX CUSTOM COMMANDS. 
\setbeamertemplate{theorems}[numbered] % ADD NUMBERING TO ALL AMS THEOREMS. 
\setbeamertemplate{footline}[frame number] % ADD PAGE NUMBERS ON BOTTOM. 
\setbeamertemplate{blocks}[rounded][shadow=true]
\setbeamertemplate{navigation symbols}{} 
% \setbeamercolor{block title}{bg=cyan,fg=black}  % CHANGE THE BLOCK STYLE IN BEAMER. 
% \setbeamercolor{block body}{bg=lime,fg=black} % CHANGE THE BLOCK STYLE IN BEAMER.

@@

\section{Introduction and background}
    \subsection{Motivations}
        \begin{frame}{Nesterov's accelerated gradient in 1983}
            Let's recall a classic. 
            \begin{block}{Nesterov's accelerated gradient in 1983}
                Initialized $x_0 = y_0 \in \RR^n$. 
                The original formulation by Nesterov in 1983 \cite{nesterov_method_1983} has

@@

                    & x_{k + 1} := y_k - L^{-1}\nabla F(y_k),
                    & t_{k + 1} := 1/2\left(1 + \sqrt{1 + 4t_{k}^2}\right),
                    & \theta_{k + 1} := (t_{k} - 1)/t_{k + 1}, \label{eqn:example-algorithm}
                    & y_{k + 1} := x_{k + 1} + \theta_{k + 1}(x_{k + 1} - x_k).

@@

            \end{block}
            If $F$ is convex and has $L$ Lipschitz smooth gradient and a minimizer $x^+$ exists. 
            Then it has
            \begin{itemize}
                \item a convergence rate of $\mathcal O(1/k^2)$ for $(F(x_k) - F(x^+))_{k \ge 1}$;
                \item this convergence rate is optimal in the sense as proposed by Nesterov \cite{nesterov_lectures_2018}.
            \end{itemize}

@@

        \begin{frame}{Other choices for the momentum sequence}
            Generalized by Beck \cite{beck_fast_2009-1} in 2009, called FISTA, it has $(t_k)_{k}$: 
            \begin{tcolorbox}\noindent\vspace{-1em}

@@

                    (\forall k \ge 1)\quad t_k (t_k - 1) &\le t_{k - 1}^2. 
            \end{tcolorbox}
            Chambolle and Dossal in 2015 \cite{chambolle_convergence_2015}, if $t_k = (k + a -1)/a$, then$(x_k)_{k \ge 0}$ has weak convergence in Hilbert space for all $a > 2$.
            \par  
            In Nesterov's book \cite{nesterov_lectures_2018} and Beck's book \cite{beck_first-order_2017} that if in addition, $F:\RR^n \rightarrow \RR$ is $L$ smooth gradient, with Strongly convex with $\mu > 0$, then $(\theta_k)_{k\ge 0}$: 
            \begin{tcolorbox}
                   \theta_k = \frac{\sqrt{L/\mu} - 1}{\sqrt{L/\mu} + 1}. 
            \end{tcolorbox}
            It gives $\mathcal O((1 - \sqrt{\mu/L})^k)$ convergence rate for $(F(x_k) - F(x^+))_{k \ge 1}$. 
    \subsection{Our discoveries and inspirations}
        \begin{frame}{A major question}
            This is a question that lead our investigation: 
            \begin{tcolorbox}
                How much can we relax the choice of the sequence $\theta_k$ and still get an upper bound for the convergence rate? 
            \end{tcolorbox}
        \begin{frame}{The assumptions we make throughout}
            \begin{assumption}[Convex smooth plus nonsmooth]
                Let the ambient space be $\RR^n$, equipped with Euclidean inner product and norm. 
                Define $F := f + g$.
                    \item $f: \RR^n \rightarrow \RR$ is $L$ Lipschitz smooth and $\mu \ge 0$ strongly convex.
                    \item $g: \RR^n \rightarrow \overline \RR$ is proper, closed and convex. The extended real is defined as $\overline \RR := \RR \cup \{\infty\}$.
                    \item A minimizer exists for the optimization problem: $F^+ = \min_x \left\lbrace f(x) + g(x)\right\rbrace$.
            \end{assumption}
            \textbf{This assumption the full scope of the theoretical and practical discussion, assume it throughout.}
        \begin{frame}{Theoretical results summary}
            Let $(\alpha_k)_{k \ge0}, (\rho_k)_{k \ge 0}$ be two sequences. 
            Suppose that $\alpha_0 \in (0, 1]$ and for all $k \ge 1$, $\alpha_k \in (\mu/L, 1)$ and define $(\rho_k)_{k\ge0 }$ to be: 
            \begin{tcolorbox}\noindent\vspace{-1em}
                    \rho_k &:= \frac{\alpha_{k + 1}^2 - (\mu/L)\alpha_{k + 1}}{(1 - \alpha_{k + 1})\alpha_k^2} \quad \forall (k \ge 0).
            \end{tcolorbox}
            We can show Nesterov's type accelerated proximal gradient algorithm (or equivalently, FISTA with $g \equiv 0$) generates $(F(x_k) - F^+)_{k\ge 1}$ that has an upper bound of: 
            \begin{tcolorbox}\noindent\vspace{-1em} 
                    \mathcal O\left(

@@

            \end{tcolorbox}

@@

        \begin{frame}{A practical realization inspired by the theories}
            Our theories suggest that, the upperbound for function value in the R-WAPG algorithm is ultimately described by the sequence $(\alpha_k)_{k \ge 0}$.
            \textbf{Taking this idea further, this is what we did for numerical experiments}: 
            \begin{tcolorbox}
                We made a Free R-WAPG algorithm which estimates the strong convexity constant $\mu$ through the Bregman Divergence at two successive iterates while the algorithm is running and, we check the upper bound using our convergence theory. 
            \end{tcolorbox}

\section{The R-WAPG framework and convergence rate}
    \subsection{Introducing the R-WAPG framework}
        \begin{frame}{Proximal gradient, gradient mapping}
            \begin{definition}[The proximal gradient operator]
                Define for all $x\in \RR^n$: 
                    T_L(y) 
                    &:= \argmin_{x \in \RR^n} \left\lbrace
                        g(x) + \langle \nabla f(y), x\rangle + L/2\Vert x - y\Vert^2
                    \right\rbrace 
                    &= \left(I + L^{-1}\partial g\right)^{-1}
                    \left(I - L^{-1}\nabla f\right)(y),
                    \mathcal G_L(y)
                    &:= L(I - T_L)y.
            \end{definition}
            Note, the $I$ here is the identity operator in $\RR^n$. 
            $L > 0$ is the Lipschitz smoothness parameter for $f$. 
        \begin{frame}{The R-WAPG sequences}
            In this section, we highlight key theoretical results that allowed us to unify convergence rate for various variants of FISTA in the literature.
            \begin{definition}{R-WAPG sequences}\label{def:rwapg-seq}
                Assume $0 \le \mu < L$. 
                Let $(\alpha_k)_{k \ge 0}$ be such that: 
                    & \alpha_0 \in (0, 1], \\
                    & \alpha_k \in (\mu/L, 1) \quad (\forall k \ge 1). 
                Define $(\rho_k)_{k \ge 0}$
                    \rho_k := \frac{\alpha_{k + 1}^2 - (\mu/L) \alpha_{k + 1}}{
                        (1 - \alpha_{k + 1}) \alpha_k^2.
                We call the sequence $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge0}$ the R-WAPG sequences. 
            \end{definition}
        \begin{frame}{R-WAPG}
            \begin{definition}[relaxed weak accelerated proximal gradient (R-WAPG)]\label{def:wapg}
                Choose any $x_1 \in \RR^n, v_1 \in \RR^n$.
                Let $(\alpha_k)_{k \ge0}, (\rho_k)_{k \ge 0}$ be given by Definition \ref{def:rwapg-seq}.
                The algorithm generates a sequence of vector $(y_k, x_{k + 1}, v_{k + 1})_{k \ge 1}$ for all $k\ge 1$ by such that:
                \begin{tcolorbox}\vspace{-1em}
                        \gamma_k &\defeq \rho_{k -1}L\alpha_{k - 1}^2,
                        L\alpha_k^2 &= (1 - \alpha_k)\gamma_k + \mu \alpha_k, 
                        \hat \gamma_{k + 1} & \defeq L\alpha_k^2,
                        y_k &=
                        (\gamma_k + \alpha_k \mu)^{-1}(\alpha_k \gamma_k v_k + \hat\gamma_{k + 1} x_k),
                        g_k &= \mathcal G_L (y_k),
                        v_{k + 1} &=
                        \hat\gamma^{-1}_{k + 1}
                        (\gamma_k(1 - \alpha_k) v_k - \alpha_k g_k + \mu \alpha_k y_k),
                        x_{k + 1} &= T_L y_k.
                \end{tcolorbox}
            \end{definition}
    \subsection{Different representation of accelerated proximal gradient}
                then the iterates can be expressed without $(\gamma_k)_{k \ge1},(\hat \gamma_k)_{k \ge 2}$, and they satisfy for all $k\ge 1$: 
    \subsection{Unifying different variants of accelerated proximal gradient}
        \begin{frame}{Convergence of Chambolle and Dossal variant of FISTA}
            FISTA algorithm proposed in Chambolle and Dossal \cite{chambolle_convergence_2015} has a function value convergence fully described by Proposition \ref{prop:wapg-convergence}. 
            \begin{lemma}[R-WAPG sequence as inverted FISTA sequence]\label{lemma:inverted-fista-seq}
                Let R-WAPG sequence $(\rho_k)_{k \ge 0}, (\alpha_k)_{k \ge 0}$ be given by Definition \ref{def:rwapg-seq}.
                If $\mu = 0, \rho_k \ge 1\; \forall k \ge 0$, and $\alpha_0 = 1$, then:
                    \item $\alpha_k^{-2} \ge \alpha_{k + 1}^{-2} - \alpha_{k + 1}^{-1}\; \forall k \ge 0$.
                    \item Let $t_k := \alpha_k^{-1}$, then $0 < t_{k + 1} \le (1/2)\left(1 + \sqrt{1 + 4t_k^2}\right)\;\forall k\ge 0$, hence the name: ``inverted FISTA sequence''.
                    \item $\prod_{i = 1}^k\max(1, \rho_{k - 1})(1 - \alpha_k) = \alpha_k^2 \quad (\forall k \ge 1)$.
            \end{lemma}
            Let $\alpha_k = a/(k + 1)$, $a \ge 2$, Applying Proposition \ref{prop:wapg-convergence} yields $\mathcal O(1/k^2)$ convergence rate of $(F(x_k) - F^+)_{k \ge 1}$ after some algebra routine. 
        \begin{frame}{In addition, our theory made a new prediction}
            \begin{theorem}[fixed momentum APG]\label{thm:fixed-momentum-fista}
                Assume $L > \mu > 0$, let a pair of constant R-WAPG sequence: $(\alpha_k)_{k \ge0}$ is a constant and $\alpha_k \in (\mu/L, 1)$ for all $k \ge 0$.
                Define $q := \mu/L$ and for any fixed $r \in \left(\sqrt{q}, \sqrt{q^{-1}}\right)$, represent $\alpha_k = \alpha = r \sqrt{q}$. 
                Consider the algorithm with a constant momentum specified by the following:
                \begin{tcolorbox}
                    Define $\theta = \left(1 - r^{-1}\sqrt{q}\right)(1 - r\sqrt{q})(1 - q)^{-1}$.
                    Initialize $y_1 = x_1$; for $k = 1, 2, \ldots, N$, update:
                        &x_{k + 1} = y_k + L^{-1}\mathcal G_L (y_k)
                        ,
                        & y_{k + 1} = x_{k + 1} + \theta(x_{k + 1} - x_k).
                \end{tcolorbox}
                Then the algorithm generates $(x_k)_{k \ge 1}$ such that $(F(x_{k}) - F(x^*))_{k\geq 1}$ converges at a rate of $\mathcal O\left(\left(1 - \min\left(\mu/(\alpha L), \alpha\right)\right)^k\right)$.
            \end{theorem}

@@

    \subsection{Free R-WAPG}

@@

    \subsection{Numerical experiment, simple quadratic}

@@

    \subsection{Numerical experiment, LASSO}
        \begin{frame}{The LASSO problem}
            The problem of LASSO from Tibshirani \cite{tibshirani_regression_1996} is the optimization problem $\min_{x \in \RR^n}\{(1/2)\Vert Ax - b\Vert^2 - \lambda\Vert x\Vert_1\}$. 
                \item $A \in \RR^{N \times N}$, full of i.i.d random variable from a standard normal distribution. 
                \item Computed in prior, $L, \mu$ are parameters estimated by $\mu = 1/\Vert (A^TA)^{-1}\Vert$ and $L = \Vert A^TA\Vert$. The norm is the spectral norm. 
                \item The smallest value of $F(x_k)$, for all algorithms, over all tries are taken as an estimate of $F^+$. 
        \begin{frame}{Results plotted}
            With $A\in \RR^{N\times N}$ fixed, for each $x_0 \sim \mathcal N(I, \mathbf 0)$ realized, 30 experiments are preformed and the medium, minimum and maximum of $\delta_k$ are recorded for M-FISTA, Free R-WAPG, and V-FISTA. 
                    \includegraphics[width=\textwidth]{assets/lasso_batched_statistics_64-256.png}
                    \caption{LASSO experiment with $M = 64, N = 256$. Plots of minimum, maximum, and median $\delta_k$ with estimated $F^*$. }
                    \includegraphics[width=\textwidth]{assets/lasso_batched_statistics_64-128.png}
                    \caption{LASSO experiment with $M = 64, N = 128$. Plots of minimum, maximum, and median $\delta_k$ with estimated $F^*$. }
                \caption{LASSO experiments statistics for test algorithms. }
                \label{fig:batched-lasso}
        \begin{frame}{Estimates of strong convexity constant}
            The parameters in this specific trial are $\mu = 7.432363627613958\times 10^{-18}$ and $L = 2321.737206983643$. 
                    \includegraphics[width=\textwidth]{assets/lasso_loss_256.png}
                    \caption
                    {A single run of LASSO experiment displaying $F(x_k) - F^*$ for several test algorithms.
                    \includegraphics[width=\textwidth]{assets/lasso_sc_estimates_256.png}
                    \caption{The $\mu_k$ estimated by FR-WAPG for one LASSO experiment. }
                \caption{A single LASSO experiment results, with $M = 64, N=256$.}
                \label{fig:single-lass-mu-estimates}
            Below is the plot of the real time estimates of R-WAPG convergence bound using for the LASSO experiment. 
                        assets/lasso_rwapg_upperbnd_256.png
                \label{fig:single-lass-r-wapg-rwapg-upperbnd}

>>>>

% THEMES AND BEAMER SETTINGS ===================================================
% \usetheme{Madrid}
\graphicspath{{.}}
\input{presets/wang/custom_commands.tex} % IMPORT WANG'S LATEX CUSTOM COMMANDS. 
\setbeamertemplate{theorems}[numbered] % ADD NUMBERING TO ALL AMS THEOREMS. 
\setbeamertemplate{footline}[frame number] % ADD PAGE NUMBERS ON BOTTOM. 
\setbeamertemplate{blocks}[rounded][shadow=true]
\setbeamertemplate{navigation symbols}{} 
\setbeamertemplate{itemize subitem}{$\circ$}
% \setbeamercolor{block title}{bg=cyan,fg=black}  % CHANGE THE BLOCK STYLE IN BEAMER. 
% \setbeamercolor{block body}{bg=lime,fg=black} % CHANGE THE BLOCK STYLE IN BEAMER.

@@

\section{Introduction and background}
    \begin{frame}{Nesterov's accelerated gradient in 1983}
        \begin{block}{Nesterov's accelerated gradient in 1983}
            Initialized $x_0 = y_0 \in \RR^n$. 
            The original formulation by Nesterov in 1983 \cite{nesterov_method_1983} has
            {\small
            \begin{align*}
                & x_{k + 1} := y_k - L^{-1}\nabla F(y_k),
                \\
                & t_{k + 1} := 1/2\left(1 + \sqrt{1 + 4t_{k}^2}\right),
                \\
                & \theta_{k + 1} := (t_{k} - 1)/t_{k + 1}, \label{eqn:example-algorithm}
                \\
                & y_{k + 1} := x_{k + 1} + \theta_{k + 1}(x_{k + 1} - x_k).
            \end{align*}    
            }
        \end{block}
        When $F$ is convex and, $L$ Lipschitz smooth. 
        If minimizer $x^+$ exists, then: 
        Then it has
        \begin{itemize}
            \item a convergence rate of $\mathcal O(1/k^2)$ for $(F(x_k) - F(x^+))_{k \ge 1}$;
            \item this convergence rate is optimal in the sense as proposed by Nesterov \cite{nesterov_lectures_2018}.
        \end{itemize}
    \end{frame}
    \begin{frame}{Choices of momentum parameters $t_k, \theta_k$}
        \begin{block}{Here are prominent results in the literature}
            \begin{itemize}
                \item Beck, Teboulle 2009 \cite{beck_fast_2009}, the FISTA paper: $t_k (t_k - 1) \le t_{k - 1}^2$. 
                    \begin{itemize}
                        \item $(F(x_k) - F(x^+))_{k \ge 0}$ converges at rate $\mathcal O(1/k^2)$ when $t_k(t_k - 1) = t_{k - 1}^2$. 
                        \item $\theta_k$ admits no closed form. 
                    \end{itemize}
                \item Chambolle and Dossal 2015 \cite{chambolle_convergence_2015}: $t_k = (k + a -1)/a$ with $a > 2$. 
                    \begin{itemize}
                        \item Retains convergence rate $\mathcal O(1/k^2)$ with $\theta_{k} = \frac{k - 2}{k + a - 1}$. 
                        \item Iterates has weak convergence in Hilbert space. 
                    \end{itemize}
                \item Nesterov \cite{nesterov_lectures_2018} and, Beck \cite{beck_first-order_2017}
                : $\theta_k = \frac{\sqrt{L/\mu} - 1}{\sqrt{L/\mu} + 1}$.
                    \begin{itemize}
                        \item Converges at rate $\mathcal O\left(\left(1 - \sqrt{\mu/L}\right)^k\right)$, if $f$ is also $\mu$ strongly convex with $\mu > 0$. 
                    \end{itemize}
                \item Apidopoulos et al. 2018 \cite{apidopoulos_convergence_2018}: $\theta_k = \frac{n}{n + b}$ for $b \in (0, 3)$. 
                \begin{itemize}
                    \item Has convergence rate $\mathcal O(1/k^{2b/3})$. 
                \end{itemize}
            \end{itemize}    
        \end{block}
    \end{frame}
    \begin{frame}{A major question}
        The following question lead our investigation: 
        \begin{tcolorbox}
            How much can we relax the choice of $\theta_k, t_k$ and still get an upper bound for the convergence rate? 
        \end{tcolorbox}
        We proposed a framework called Relaxed, Weak Accelerated Proximal Gradient (R-WAPG) and, these are what we did: 
        \begin{itemize}
            \item An upper bound of function value with the weakest assumption on momentum parameter.
            \item A unified analysis with strong convexity $\mu \ge 0$. 
            \item Various equivalent representation of R-WAPG are given and, convergence results varified for examples in the literature. 
            \item Numerical experiments are conducted to verify the theory. 
        \end{itemize}
    \end{frame}

\section{The R-WAPG Framework}
    \begin{frame}{The assumptions we make throughout}
        \begin{assumption}[Convex smooth plus nonsmooth]
            Let the ambient space be $\RR^n$, equipped with Euclidean inner product and norm. 
            Define $F := f + g$.
            \begin{enumerate}
                \item $f: \RR^n \rightarrow \RR$ is $L$ Lipschitz smooth and $\mu \ge 0$ strongly convex.
                \item $g: \RR^n \rightarrow \overline \RR$ is proper, closed and convex. The extended real is defined as $\overline \RR := \RR \cup \{\infty\}$.
                \item A minimizer exists for the optimization problem: $F^+ = \min_x \left\lbrace f(x) + g(x)\right\rbrace$.
            \end{enumerate}
        \end{assumption}
        \textbf{This assumption the full scope of the theoretical and practical discussion, assume it throughout.}
    \end{frame}
    \begin{frame}{Proximal gradient, gradient mapping}
        \begin{definition}[The proximal gradient operator]
            Define for all $x\in \RR^n$: 
            \begin{align*}
                T_L(y) 
                &:= \argmin_{x \in \RR^n} \left\lbrace
                    g(x) + \langle \nabla f(y), x\rangle + L/2\Vert x - y\Vert^2
                \right\rbrace 
                \\
                &= \left(I + L^{-1}\partial g\right)^{-1}
                \left(I - L^{-1}\nabla f\right)(y),
                \\
                \mathcal G_L(y)
                &:= L(I - T_L)y.
            \end{align*}
        \end{definition}
        Note, the $I$ here is the identity operator in $\RR^n$. 
        $L > 0$ is the Lipschitz smoothness parameter for $f$. 
    \end{frame}
    \begin{frame}{The R-WAPG sequences}
        In this section, we highlight key theoretical results that allowed us to unify convergence rate for various variants of FISTA in the literature.
        \begin{definition}[R-WAPG sequences]\label{def:rwapg-seq}
            Assume $0 \le \mu < L$. 
            Let $(\alpha_k)_{k \ge 0}$ be such that: 
            \begin{align*}
                & \alpha_0 \in (0, 1], \\
                & \alpha_k \in (\mu/L, 1) \quad (\forall k \ge 1). 
            \end{align*}    
            Define $(\rho_k)_{k \ge 0}$
            \begin{align*}
                \rho_k := \frac{\alpha_{k + 1}^2 - (\mu/L) \alpha_{k + 1}}{
                    (1 - \alpha_{k + 1}) \alpha_k^2.
                }
            \end{align*}
            We call the sequence $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge0}$ the R-WAPG sequences. 
        \end{definition}
    \end{frame}
    \begin{frame}{R-WAPG}
        \begin{definition}[R-WAPG]\label{def:wapg}
            Choose any $x_1 \in \RR^n, v_1 \in \RR^n$.
            Let $(\alpha_k)_{k \ge0}, (\rho_k)_{k \ge 0}$ be given by Definition \ref{def:rwapg-seq}.
            The algorithm generates a sequence of vector $(y_k, x_{k + 1}, v_{k + 1})_{k \ge 1}$ for all $k\ge 1$ by such that:
            \begin{tcolorbox}\vspace{-1em}
                \begin{align*}
                    \gamma_k &\defeq \rho_{k -1}L\alpha_{k - 1}^2,
                    \\
                    L\alpha_k^2 &= (1 - \alpha_k)\gamma_k + \mu \alpha_k, 
                    \\
                    \hat \gamma_{k + 1} & \defeq L\alpha_k^2,
                    \\
                    y_k &=
                    (\gamma_k + \alpha_k \mu)^{-1}(\alpha_k \gamma_k v_k + \hat\gamma_{k + 1} x_k),
                    \\
                    g_k &= \mathcal G_L (y_k),
                    \\
                    v_{k + 1} &=
                    \hat\gamma^{-1}_{k + 1}
                    (\gamma_k(1 - \alpha_k) v_k - \alpha_k g_k + \mu \alpha_k y_k),
                    \\
                    x_{k + 1} &= T_L y_k.
                \end{align*}
            \end{tcolorbox}
        \end{definition}
    \end{frame}

\section{Our Theoretical Results}
            Then they satisfy for all $k\ge 1$: 

@@

@@

@@

@@

@@

@@

@@

@@

@@
